---
title: "Visualizing NYC Dog Data"
author: "Kelli-Jean Chun"
date: "11/04/2018"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.align = 'center')

# Load packages
library("pacman")
p_load("zipcode", "tidyverse",
       "rgdal", "rgeos",
       "maptools", "ggalt",
       "ggthemes", "ggrepel",
       "RColorBrewer", "mapproj",
       "scales", "cowplot", "knitr",
       "kableExtra", 
       "reticulate" # connect Python and R
       )

# Load helpers
source("helpers.R")
```


```{r load_data, include=FALSE}
# Dog data - found dataset through Data is Plural
# https://tinyletter.com/data-is-plural/letters/data-is-plural-2017-05-31-edition
# https://project.wnyc.org/dogs-of-nyc/
snapshot_date = as.Date("2013-01-24")
dogs = read.csv("data/Dogs of NYC _ WNYC.csv")
```

## Raw Data

Below is a subset of the raw NYC dogs data, which can be found [here](https://project.wnyc.org/dogs-of-nyc/).

```{r display_data}
set.seed(123)
kable(head(dogs[sample(1:nrow(dogs)), ]), row.names=FALSE) %>%
  kable_styling("hover") %>%
  scroll_box(width = "100%", height = "300px")
```


```{r process_data}
# Process Birth Year and Month
dogs$birth_year = substr(dogs$birth, 5, 6)
dogs$birth_month = substr(dogs$birth, 1 ,3)
dogs$birth_year[dogs$birth_year == ""] = NA
dogs$birth_month = ifelse(is.na(dogs$birth_year), NA, dogs$birth_month)
dogs$birth_clean = as.Date(ifelse(!is.na(dogs$birth_year), 
                                  paste0(dogs$birth_year, "-", 
                                         dogs$birth_month, 
                                         "-01"),
                                  NA),
                           format="%y-%b-%d")

dogs$is_birth_invalid = ifelse((dogs$birth_clean > snapshot_date)  | 
                                 (is.na(dogs$birth_year)), 
                               TRUE, 
                               FALSE)
# Filter out dogs with invalid birth years
dogs = dogs[!dogs$is_birth_invalid, ] 

# Days since the dog was born
dogs$days_since_birth = as.numeric(snapshot_date - dogs$birth_clean)
age_95th = quantile(dogs$days_since_birth, 0.95)
# Remove dogs above the 95th percentile, they're probably dead :(
dogs = dogs[dogs$days_since_birth < age_95th, ]

# Process dog colors
dogs$dominant_color = as.factor(dogs$dominant_color)
dogs$secondary_color[dogs$secondary_color == "n/a"] = NA
dogs$secondary_color = as.factor(dogs$secondary_color)
dogs$third_color[dogs$third_color == "n/a"] = NA
dogs$third_color = as.factor(dogs$third_color)

dogs$spayed_or_neutered_bool = dogs$spayed_or_neutered == "Yes"
dogs$zip_group = group_zip_code(dogs$zip_code)

# Process categorical
dogs$is_born_jan = dogs$birth_month == "Jan"
dogs$is_male = dogs$gender == "M"
#dogs$dog_name_clean = tools::toTitleCase(tolower(dogs$dog_name))
dogs$dog_name_clean = dogs$dog_name
dogs$dog_name_clean = gsub(" " , "", dogs$dog_name_clean)
dogs$dog_name_clean = ifelse(dogs$dog_name_clean == 'n/a', 
                             "N/A", dogs$dog_name_clean)

dogs$is_name_missing = dogs$dog_name_clean == "N/A"
dogs$is_pit_bull = dogs$breed == "American Pit Bull Terrier/Pit Bull"

n_male = sum(dogs$gender == "M")
# Make default gender the most common
default_gender = ifelse(n_male >= nrow(dogs[dogs$gender != 'n/a',])/2, "M", "F")
dogs$gender_clean = ifelse(dogs$gender == 'n/a', default_gender, dogs$gender)

dogs$is_secondary_color_missing = is.na(dogs$secondary_color)
dogs$is_third_color_missing = is.na(dogs$third_color)

# Data frame to be read into Python for modelling
# Collect features  -------------------------------------------------------
response = "spayed_or_neutered_bool"
features = c("days_since_birth", "is_name_missing", "is_pit_bull", "is_male", 
             "borough", "is_born_jan", "is_secondary_color_missing", "is_third_color_missing")
df = dogs[, c(features, response)]

saveRDS(df, "data/dogs_proc.RDS")
```



## Days since birth

Dogs that are older are more likely to be spayed/neutered.


```{r days_since_birth}
ggplot(dogs, aes(y=days_since_birth, x=spayed_or_neutered, fill=spayed_or_neutered)) + 
  geom_boxplot() + 
  theme_classic() + 
  ylab("Days Since Birth") + 
  xlab("Spayed/Neutered") + 
  scale_y_continuous(labels=comma) + 
  scale_fill_brewer(palette = "Set1") + 
  theme(legend.position="none")

```

```{r geography, include=FALSE}
# zip code shape files from:
# https://www.census.gov/geo/maps-data/data/cbf/cbf_zcta.html
# Code adjusted from this stack overflow answer: https://stackoverflow.com/questions/47602744/draw-boundary-by-zip-code-and-create-a-heat-map

# Load the zip poly data for US
map_data = readOGR(dsn = "data/cb_2017_us_zcta510_500k", 
                   layer = "cb_2017_us_zcta510_500k")
# Get polygon data for dog data only
map_poly = subset(map_data, ZCTA5CE10 %in% dogs$zip_code)

# Create a new group with the first three digit.
# Drop unnecessary factor levels.
# Add a fake numeric variable, which is used for coloring polygons later.
map_poly$group = group_zip_code(map_poly$ZCTA5CE10)

# Merge polygons using the group variable
# Create a data frame for ggplot.
map_poly.union = unionSpatialPolygons(map_poly, map_poly$group)
map_data = fortify(map_poly.union)

# Convert SpatialPolygons to data frame and aggregate the fake values
map_poly.df = as(map_poly, "data.frame")
map_poly.df = merge(map_poly.df, dogs, 
                    by.x="ZCTA5CE10", by.y="zip_code", all.y=TRUE)
map_poly.df$group[is.na(map_poly.df$group)] = 
  group_zip_code(map_poly.df$ZCTA5CE10[is.na(map_poly.df$group)])
map_poly.df = map_poly.df %>%
  group_by(group, borough) %>%
  summarise(value = mean(spayed_or_neutered_bool))
map_poly.df$borough_zip = paste0(map_poly.df$borough,
                                 " (", map_poly.df$group, ")")

# Find a center point for each zip code area
centers = data.frame(gCentroid(spgeom = map_poly.union, byid = TRUE))
centers$zip = rownames(centers)
centers = merge(centers, map_poly.df, by.x="zip", by.y="group")

```

## Geographical differences 

```{r plot_geography, echo=TRUE, fig.cap='Spayed/Neutered Rate for NYC Boroughs and Zips'}

# Zip - Borough plot
ggplot() +
  geom_cartogram(data = map_data, 
                 aes(x = long, y = lat, map_id = id), 
                 map = map_data) +
  geom_cartogram(data = map_poly.df, 
                 aes(fill = value, map_id = group), 
                 map = map_data) +
  geom_text_repel(data = centers, 
                  aes(label = borough_zip, x = x, y = y), size = 3) +
  scale_fill_gradientn(colours = (brewer.pal(5, "Spectral"))) +
  coord_map() +
  theme_map() +
  labs(fill="Spayed/Neutered Rate\n") + 
  theme(legend.position="top") 

```



## Categorical features

### Top dog names
```{r name, fig.width=8, fig.height=4}
by_name = plot_categorical(dogs, "dog_name_clean", "Top Dog Names")
by_name$plot
```

### Top Breeds

```{r breed, fig.width=8, fig.height=4}
by_breed = plot_categorical(dogs, "breed", "Top Breeds")
by_breed$plot
```



### Gender
```{r gender, fig.width=8, fig.height=4}
by_gender = plot_categorical(dogs, "gender_clean", "By Dog Gender", ylimits=c(0.7, 0.85))
by_gender$plot

```


### Colors
```{r colors, fig.width=8, fig.height=4}
by_dominant_color = plot_categorical(dogs, "dominant_color", "Top Dominant Colors")
by_dominant_color$plot


by_second_color = plot_categorical(dogs, "secondary_color", "Top Secondary Colors")
by_second_color$plot

by_third_color = plot_categorical(dogs, "third_color", "Top Third Colors")
by_third_color$plot

```

### Guard/Trained

```{r, fig.width=8, fig.height=4}
by_trained = plot_categorical(dogs, "guard_or_trained", "By Guard/Trained",
                              ylimits=c(0.45, 0.8))
by_trained$plot
```

### Birth Month

```{r, fig.width=8, fig.height=4}
by_month = plot_categorical(dogs, "birth_month", "By Birth Month",
                              n=13,
                            factor_order = rev(month.abb))
by_month$plot
```


## Processed Data 
Data to be used to train a model in Python. 

```{r display_processed_data}
kable(head(df[sample(1:nrow(df)), ]), row.names=FALSE) %>%
  kable_styling("hover") %>%
  scroll_box(width = "100%", height = "300px")
```

## Python in R

An example of connecting Python and R

```{python, echo=TRUE}
print('Python in R')
for i in range(3):
  print(i)
```


### Running Jupyter in Python

Several ways to execute a Jupyter notebook in R: 

1. Run Python in R and execute via papermill:

```{python modelling, results='hide', echo=TRUE}
import papermill as pm
pm.execute_notebook("example_notebook.ipynb", "executed_notebook/example_notebook.ipynb")
```


2. In R, invoke terminal and execute with nbconvert:

```{r jupyter_example, results='hide', echo=TRUE}
system(command="jupyter nbconvert --ExecutePreprocessor.timeout=600 --to notebook --execute example_notebook.ipynb")
```

### Predicting spayed/neutered

The selected model based on AUC is a GBM:

```{r}
scores = read.csv("python_output/gbm_scores.csv", check.names=FALSE)
feat_imp = read.csv("python_output/gbm_feat_imp.csv", header=FALSE)

scores = data.frame(t(apply(scores, 2, round, digits=3)), check.names=FALSE)
names(scores)[names(scores) == "Avg CV Score"] = "Avg"
kable(scores) %>%
  kable_styling("hover", full_width = FALSE, position = "center") %>%
  add_header_above(c("CV Score" = 4, " " = 1))

p_load("formattable")
names(feat_imp) = c("Feature", "Importance")
feat_imp$Feature = gsub("_", " ", feat_imp$Feature)
feat_imp$Importance = round(feat_imp$Importance, 2)
feat_imp %>%
  mutate(Importance = color_bar("lightgreen")(Importance)) %>%
  kable(escape = FALSE) %>%
  kable_styling("hover", full_width = FALSE, position = 'center') %>%
  column_spec(1, width = "7cm")

```



